---
title: "Fitness Dynamics D68"
#author: "David Jorgensen"
#date: "`r Sys.Date()`"
output: html_document
---
Here we apply the fitness dynamic method of LeFrancq et al. to publicly available EV-D68 data.

```{r necessary packages, echo = T, eval=T, , results = 'hide', warning=FALSE, message=FALSE}
# Required R packages
pacman::p_load(ape,        
               phytools,
               stringr,
               MetBrewer,
               parallel,
               mgcv,
               cowplot,
               ggplot2,
               ggtree,
               devtools,
               lubridate,
               dplyr)
devtools::install_github("rasigadelab/thd")
library(thd)
```

```{r functions from noemie repo, eval=T, results = 'hide', warning=FALSE, message=FALSE}
# from cloned repo
#source(file = '../../paper-index-fitness-dynamics-trees/2_Functions/2_1_Index_computation_20231220.R')
#source(file = '../../paper-index-fitness-dynamics-trees/2_Functions/2_2_Lineage_detection_20231220.R')

# direct from github
source_url("https://raw.githubusercontent.com/noemielefrancq/paper-index-fitness-dynamics-trees/main/2_Functions/2_1_Index_computation_20231220.R")
source_url("https://raw.githubusercontent.com/noemielefrancq/paper-index-fitness-dynamics-trees/main/2_Functions/2_2_Lineage_detection_20231220.R")
```

Tree downloaded from nextstrain 07/02/24. Metadata added to tip labels.
```{r load data, eval=T, results = 'hide', warning=FALSE, message=FALSE}
## Tree with location, date and lineage at tips - extract this data after loading - will try with full genome and all data first
D68_WGS_tree_dated <- read.nexus("nextstrain_enterovirus_d68_genome_timetree.nexus")
## Make sure the tree is binary, and ladderized
D68_WGS_tree_dated <- collapse.singles(ladderize(multi2di(D68_WGS_tree_dated, random = F), right = F))

## Load metadata
D68_meta <- read.delim("nextstrain_enterovirus_d68_genome_metadata (2).tsv", sep = "\t")

## Generate new tip names and order (Format: virus/Location/seqname/decimal date/clade)
D68_meta$newname <- paste("EV-D68",D68_meta$country, D68_meta$accession, round(decimal_date(as.Date(word(D68_meta$date, 1))),3), D68_meta$clade_membership,   sep = "/")

## Assign to tree
names_seqs <- D68_WGS_tree_dated$tip.label
D68_meta <- D68_meta[match(names_seqs, D68_meta$strain),]
D68_WGS_tree_dated$tip.label <- D68_meta$newname 

## Save tree
write.nexus(D68_WGS_tree_dated, file = "reformat_tree.nexus", translate = F)

## Extract these params from the tip names (useful for homemade trees)
names_seqs <- D68_WGS_tree_dated$tip.label
n_seq = length(names_seqs)
## Collection times of all sequences
times_seqs = as.numeric(sapply(names_seqs, function(x)tail(str_split(x, pattern = '/')[[1]],2)[1]))
## Nextstrain clades of all sequences
clades_seqs = sapply(names_seqs, function(x)tail(str_split(x, pattern = '/')[[1]],1))

```

#### Index parameters 

```{r, eval=T}
## Length genome 
genome_length = 7255
## Mutation rate 
mutation_rate = 4e-3 # clock rate from nextstrain https://nextstrain.org/enterovirus/d68/genome?l=clock

## Parameters for the index
## Window of time on which to search for samples in the population - try longer than SARS-CoV2 as not many seqs. and covers a longer time period
wind = 20*365 #days
wind = wind/365

## Calculate the timescale?
timescale = compute.timescale(genome_length, mutation_rate, wind)
print(timescale)

wind = 1
timescale = 1
```

## Compute the index dynamics

#### Compute pairwise distance matrix
Compute distance between each pair of sequences and internal nodes in the tree
```{r, eval=T}
genetic_distance_mat <- dist.nodes.with.names(D68_WGS_tree_dated)
```

Get the time of each internal node
```{r, eval=T}
nroot = length(D68_WGS_tree_dated$tip.label) + 1 ## Root number
distance_to_root = genetic_distance_mat[nroot,]
root_height = times_seqs[which(names_seqs == names(distance_to_root[1]))] - distance_to_root[1]
nodes_height = root_height + distance_to_root[n_seq+(1:(n_seq-1))]
```

#### Preparation data tips and nodes

Prepare the main dataframe, where the index and lineages of all nodes (internal and terminal) are going to be stored.
```{r, eval=T}
# Meta-data with all nodes 
dataset_with_nodes = data.frame('ID' = c(1:n_seq, n_seq+(1:(n_seq-1))),
                                'name_seq' = c(names_seqs, n_seq+(1:(n_seq-1))),
                                'time' = c(times_seqs, nodes_height),
                                'is.node' = c(rep('no', n_seq), rep('yes', (n_seq-1))),
                                'Nextstrain_clade' = c(clades_seqs, rep(NA, n_seq-1)))
```


#### Compute index of every tip and node
```{r, eval=T}
dataset_with_nodes$index = compute.index(time_distance_mat = genetic_distance_mat, 
                                         timed_tree = D68_WGS_tree_dated, 
                                         time_window = wind,
                                         metadata = dataset_with_nodes, 
                                         mutation_rate = mutation_rate,
                                         timescale = timescale,
                                         genome_length = genome_length)
```

## Plot tree & index below, with colors from NextStrain clades

First, generate the color key, based on the Nextstrain clade of each sequence.
```{r, eval=T}
## Color key fro Nextstrain clades
colors_clade = met.brewer(palette_name="Cross", n=length(levels(as.factor(dataset_with_nodes$Nextstrain_clade))), type="continuous")

## Color of each node, based on the key
dataset_with_nodes$Nextstrain_clade_color = as.factor(dataset_with_nodes$Nextstrain_clade)
clade_labels = levels(dataset_with_nodes$Nextstrain_clade_color)
levels(dataset_with_nodes$Nextstrain_clade_color) = colors_clade
dataset_with_nodes$Nextstrain_clade_color = as.character(dataset_with_nodes$Nextstrain_clade_color)
```

Then plot the tree and index:
```{r, eval=T}
par(mfrow = c(2,1), oma = c(0,0,0,0), mar = c(4,4,0,0))

min_year = 2000
max_year = 2023

## Tree
plot(D68_WGS_tree_dated, show.tip.label = FALSE, 
     edge.color = 'grey', edge.width = 0.25,
     x.lim = c(min_year, max_year)-root_height)
tiplabels(pch = 16, col = dataset_with_nodes$Nextstrain_clade_color, cex = 0.3)
axisPhylo_NL(side = 1, root.time = root_height, backward = F,
             at_axis = seq(min_year, max_year, 0.5)-root_height,
             lab_axis = seq(min_year, max_year, 0.5), lwd = 0.5)
## Index
plot(dataset_with_nodes$time, 
     dataset_with_nodes$index, 
     col = adjustcolor(dataset_with_nodes$Nextstrain_clade_color, alpha.f = 1),
     bty = 'n', xlim = c(min_year, max_year), cex = 0.5,
     pch = 16, bty = 'n', ylim = c(0, 1), 
     main = paste0(''), 
     ylab = 'Index', xlab = 'Time (years)', xaxt = 'n', yaxt = 'n')
axis(2, las = 2, lwd = 0.5)
axis(1, lwd = 0.5)

# Color key
legend('topright', 
       legend = clade_labels,
       fill = colors_clade, border = colors_clade,
       cex = 0.5, bty = 'n', ncol = 5)
```

## Find clades based on index dynamics 

#### Run the lineage detection algorithm

Parameters for the detection: 
```{r, eval = T}
time_window_initial = 2030;
time_window_increment = 100;
p_value_smooth = 0.05
weight_by_time = 0.1
k_smooth = -1
plot_screening = F
min_descendants_per_tested_node = 30
min_group_size = 30
weighting_transformation = c('inv_sqrt')

parallelize_code = T
number_cores = 6

max_stepwise_deviance_explained_threshold = 0
max_groups_found = 6
stepwise_AIC_threshold = 0

keep_track = T
```


Run the detection function:
```{r, eval=T, results = 'hide', warning=FALSE, message=FALSE}
start_time = Sys.time()
potential_splits = find.groups.by.index.dynamics(timed_tree = D68_WGS_tree_dated,
                                                 metadata = dataset_with_nodes,
                                                 node_support = D68_WGS_tree_dated$edge.length[match((n_seq+1):(2*n_seq-1), D68_WGS_tree_dated$edge[,2])],
                                                 threshold_node_support = 1/(7255*0.004),
                                                 time_window_initial = time_window_initial,
                                                 time_window_increment = time_window_increment,
                                                 min_descendants_per_tested_node = min_descendants_per_tested_node,
                                                 min_group_size = min_group_size,
                                                 p_value_smooth = p_value_smooth,
                                                 stepwise_deviance_explained_threshold = max_stepwise_deviance_explained_threshold,
                                                 stepwise_AIC_threshold = stepwise_AIC_threshold,
                                                 weight_by_time = weight_by_time,
                                                 weighting_transformation = weighting_transformation,
                                                 k_smooth = k_smooth,
                                                 parallelize_code = parallelize_code,
                                                 number_cores = number_cores, 
                                                 plot_screening = plot_screening,
                                                 max_groups_found = max_groups_found, 
                                                 keep_track = keep_track)
end_time = Sys.time()
print(end_time - start_time)
```

Optimize the number of groups: set the minimum number of sequences per group to 30, with a minimum frequency of 1%.
```{r, eval = T}
split = merge.groups(timed_tree = D68_WGS_tree_dated, metadata = dataset_with_nodes, 
                     initial_splits = potential_splits$potential_splits, 
                     group_count_threshold = 30, group_freq_threshold = 0.01)
```

Label sequences with these new groups, and assign a color to each of them.
```{r, eval = T}
## Label sequences with new groups
dataset_with_nodes$groups = as.factor(split$groups)
## Reorder labels by time of emergence
name_groups = levels(dataset_with_nodes$groups)
time_groups_world = NULL
for(i in 1:length(name_groups)){
  time_groups_world = c(time_groups_world, min(dataset_with_nodes$time[which(dataset_with_nodes$groups == name_groups[i] &
                                                                                   dataset_with_nodes$is.node == 'no')]))
}
levels(dataset_with_nodes$groups) = match(name_groups, order(time_groups_world, decreasing = T))
dataset_with_nodes$groups = as.numeric(as.character(dataset_with_nodes$groups))
dataset_with_nodes$groups = as.factor(dataset_with_nodes$groups)
## Update names in split list
split$tip_and_nodes_groups = match(split$tip_and_nodes_groups, order(time_groups_world, decreasing = T))
names(split$tip_and_nodes_groups) = 1:length(split$tip_and_nodes_groups)
split$groups = as.factor(split$groups)
levels(split$groups) = match(name_groups, order(time_groups_world, decreasing = T))
split$groups = as.numeric(as.character(split$groups))
## Choose color palette
n_groups <- length(name_groups)
colors_groups = (met.brewer(palette_name="Cross", n=n_groups, type="continuous"))
## Color each group
dataset_with_nodes$group_color = dataset_with_nodes$groups
levels(dataset_with_nodes$group_color) = colors_groups
dataset_with_nodes$group_color = as.character(dataset_with_nodes$group_color)
```

## Plot tree & index below, with colors from index-defined groups

Plot the tree and index colored with the new groups:
```{r, eval = T}
par(mfrow = c(2,1), oma = c(0,0,0,0), mar = c(4,4,0,0))

## Tree
plot(D68_WGS_tree_dated, show.tip.label = FALSE, 
     edge.color = 'grey', edge.width = 0.25,
     x.lim = c(min_year, max_year)-root_height)
tiplabels(pch = 16, col = dataset_with_nodes$group_color, cex = 0.3)
axisPhylo_NL(side = 1, root.time = root_height, backward = F,
             at_axis = seq(min_year, max_year, 0.5)-root_height,
             lab_axis = seq(min_year, max_year, 0.5), lwd = 0.5)

## Index colored by group
plot(dataset_with_nodes$time, 
     dataset_with_nodes$index, 
     col = adjustcolor(dataset_with_nodes$group_color, alpha.f = 1),
     bty = 'n', xlim = c(min_year, max_year), cex = 0.5,
     pch = 16, bty = 'n', ylim = c(0, 1), 
     main = paste0(''), 
     ylab = 'Index', xlab = 'Time (years)', yaxt = 'n')
axis(2, las = 2)
# Color key
legend('topright', 
       legend = name_groups,
       fill = colors_groups, border = colors_groups,
       cex = 0.5, bty = 'n', ncol = 5)
```
